---
title: "Data filtering and manipulation"
author: "Shelley, Sangay and AC"
date: "02/02/2022"
output:
  pdf_document: default
  html_document: default
---
```{r}
install.packages("knitr")
install.packages("vegan")
install.packages("ggplot2")
install.packages("tidyverse")
```
*Objective:* To subset morphological data by including or excluding samples based on filters (e.g. list of characters, groupings, etc.)

*Description of script:* This is a two-step filtering processes for the downstream analysis with filtering process (1) based on the characters used and (2) based on the samples to include.

```{r setup, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(cluster) #for hierarchical clustering
library(vegan) # NMDS and other statistics
library(ggplot2)
library(tidyverse)
```
Define names for input files and directories
"indir" is the directory containing the input files
```{r filenames and directories}

indir <- "../data"
rawdata <- "rawdata.csv"
rawdatafile <- file.path(indir,rawdata)
column_metadata <- "column_metadata.csv"
column_metadatafile <- file.path(indir,column_metadata)
row_metadata <- "row_metadata.csv"
row_metadatafile <- file.path(indir,row_metadata)
```


## Documentation of data filtering and manipulation for morphological analysis

The analysis is based on three data files:
  `data/rawdata.csv`
  `data/column_metadata.csv`
  `data/row_metadata.csv`

The selection of characters for downstream analyses is based on the filtering of groups in the `data/column_metadata.csv` file.
Characters are included in a group if marked with a yes (`Y`). Then these characters are pulled in the rawdata file based on the list of selected characters in the `data/column_metadata.csv`.

```{r read rawdata}
dataset0<-read_csv(rawdatafile) %>%
  glimpse()
col_meta<-read_csv(column_metadatafile) %>%
  glimpse() #List column names to select character for downstream analysis
            #"group" names correspond to the set of characters selected for a specific analysis
            #(e.g. "group1" can correspond to NMDS analysis)
row_meta<-read_csv(row_metadatafile) %>%
  glimpse()
```

# Check integrity of rawdata headings with column metadata values
Output is two lists; 
1. values which are in metadata but not in rawdata
2. values which are in rawdata but not in metadata

If no values are returned, lists match perfectly.

```{r}

## Characters in metadata without matches in dataset
col_meta$character[!col_meta$character %in% names(dataset0)]

## Characters in dataset without matches in metadata
names(dataset0)[!names(dataset0) %in% col_meta$character]
```



# Filtering of variables (characters) to use in analysis based on the column_metadata file:
Decides which columns (characters) to use 

```{r}
include_columns<-filter(col_meta,Group1=="y")
include_columns$character
dataset_columns <- select(dataset0,include_columns$character ) %>%
  glimpse()
```

## Filtering the rows of data for specific OTUs (select which samples to use)

This is from the row_metadata file.
Dataset Isla is filtering from the isla group column of the row_metadata file.
Additional columns can be added to the row_metadata file to create new groups in a similar way.
Use semi_join to keep all rows in dataset0 that match include_rows based in value of Collection number

```{r Filter_rows}
include_rows<-filter(row_meta,Islagroup=="Y")
datasetisla<- dataset0 %>% semi_join(include_rows, by = "Collection number")
```





















