---
title: "association and clustering analysis"
author: "Helen"
date: "08/02/2022"
output: html_document
---


*Import filtered data from filtration step into the assiociation analysis, this currently doesn't work, as it needs appropriate filtered data to run in daisy*

##Association analysis


###loading packages with fancy function
```{r}
# First specify the packages of interest (add extra as needed)
packages = c("xlsx", "cluster","vegan", "ggplot2", "mclust", "factoextra")

## Now load or install&load all
package.check <- lapply(
  packages,
  FUN = function(x) {
    if (!require(x, character.only = TRUE)) {
      install.packages(x, dependencies = TRUE)
      library(x, character.only = TRUE)
    }
  }
)

```


```{r Create an association matrix}
str(dataset1)
asym.cols = c(dataset1[,22])

#daisy treats factors as nominal
#This version is specific to Helens data, what we need is to edit it to bring info about column types in the metadata into this script
#dist <- daisy(dataset1, metric = "gower", type = list(symm = c(14,19))) #generates the distance matrix


 dist <- dataset1[,-1] %>% mutate_if(is.character, as.factor) %>% #This line changes all character vectors to factors.
    daisy( metric = "gower", type = list(symm= 'Calyx; sepal abaxial surface; indumentum')) #generates the distance matrix
 
 

 str(dist)
 #, type= list(symm= asym.cols) Added for binary variables
 
 
colnames(dataset1
         )
```



#show distance matrix

```{r display the distance matrix}
# have to use unlist because the vector you made was actually a list-weird?? 
label1 <- unlist(dataset1[,1])
clust <- hclust(dist, method = "average") #runs UPGMA cluster analysis
clust$labels= label1
plot(clust)#plots dendrogram
# str(clust)
 str(label1)
class(label1)
```


```{r clustering analysis}
nmds <- metaMDS(dist, k = 2) #run NMDS analysis
# stressplot(nmds)
plot(nmds, dis="sites", type="t")
plot(nmds, dis="sites", type="n")
orditorp(nmds, "sites", label = label1)
```

* **general importance** provides importance metric by comparing nmds to the original dataset
* `envfit` runs this analysis and provides a table to identify; 
  * relation of each variable to nmds axis
  * how important a given variable is 
  * if the importance of each variable is significantly higher than would be expected by random
```{r general importance}
#Make a list specifying labels for vectors and factors
vectorlabel1 <- colnames(dataset1)
vectorlabel <- c(vectorlabel1[-1])
label2 <- list("vectors" = vectorlabel ,"factors"= label1)
#generate a table of character importance
importance <- envfit(nmds, dataset1, na.rm = TRUE)
plot(nmds, "sites")
plot(importance, labels = label2)

#  str(importance)
# class(importance$factors)


```



##Making a pretty dendrogram?

```{r}
clus <- cutree(clust, 4)
g <- split(names(clus), clus)

p <- ggtree(clust, linetype='dashed')
clades <- sapply(g, function(n) MRCA(p, n))

p <- groupClade(p, clades, group_name='subtree') + aes(color=subtree)

# d <- data.frame(label = names(clus), 
#                 cyl = datasetisla[names(clus), "cyl"])

## Dummy grouping 
d <- data.frame(label = names(clus), 
                pop = samplemeta$group1[match(names(clus),samplemeta$long.id)])

p %<+% d + 
  layout_dendrogram() + 
  geom_tippoint(aes(fill=factor(pop), x=x+.5),
                size=5, shape=21, color='black') +
  geom_tiplab(aes(label=pop), size=3, hjust=.5, color='black') +
  # geom_tiplab(angle=90, hjust=1, offset=-10, show.legend=FALSE) + 
  scale_color_brewer(palette='Set1', breaks=1:4) +
  theme_dendrogram(plot.margin=margin(6,6,80,6)) +
  theme(legend.position=c(.9, .6))
```



##Mclust clustering analysis
```{r}
isladens <- densityMclust(nmds$points)
plot(isladens, what= "BIC")   #use the what command to run the plots individually without the number choice coming up each time
plot(isladens, what = "density", type = "hdr")


#cluster analysis- G forces dimensions of choice
mc <- Mclust(nmds$points, G =4)
plot(mc, what = "BIC")
plot(mc, what = "density")

summary(mc)
plot(mc)
 
```




