---
title: "association and clustering analysis"
author: "Helen"
date: "08/02/2022"
output: html_document
---

##Association analysis


###loading packages with fancy function
```{r}
# First specify the packages of interest (add extra as needed)
packages = c("xlsx", "cluster","vegan", "ggplot2", "mclust", "factoextra")

## Now load or install&load all
package.check <- lapply(
  packages,
  FUN = function(x) {
    if (!require(x, character.only = TRUE)) {
      install.packages(x, dependencies = TRUE)
      library(x, character.only = TRUE)
    }
  }
)

```


```{r Create an association matrix}
str(dataset1)
# asym.cols = c(dataset1[,26])

#daisy treats factors as nominal
#This version is specific to Helens data, what we need is to edit it to bring info about column types in the metadata into this script
#dist <- daisy(dataset1, metric = "gower", type = list(symm = c(14,19))) #generates the distance matrix


 dist <- dataset1 %>% mutate_if(is.character, as.factor) %>% 
    daisy( metric = "gower", type = list(symm= 'Calyx; sepal abaxial surface; indumentum')) #generates the distance matrix
 
 

 str(dist)
 #, type= list(symm= asym.cols) Added for binary variables
 
 
colnames(dataset1
         )
```



#Create a hierarchical clustering dendrogram using UPGMA

```{r display the distance matrix}
# have to use unlist because the vector you made was actually a list-weird?? 
label1 <- unlist(dataset1[,2])
clust <- hclust(dist, method = "average") #runs UPGMA cluster analysis
clust$labels= label1
plot(clust)#plots dendrogram
# str(clust)
 str(label1)
class(label1)
```


```{r clustering analysis}
#It would be good to make a scree plot (number of dimensions by stress value), to help determine best number of dimensions to use
#Failing that, just go for the smallest number of dimensions with a stress value between 0.05 and 0.1 if practical, but definitely less than 0.2. 
nmds <- metaMDS(dist, k = 2) #run NMDS analysis
# wascores =FALSE, autotransform = FALSE, noshare = FALSE, Trace = TRUE
# stressplot(nmds)
# str(nmds)
nmds$stress
plot(nmds, dis="sites", type="t")
plot(nmds, dis="sites", type="n")
orditorp(nmds, "sites", label = label1)
```

* **general importance** provides importance metric by comparing nmds to the original dataset
* `envfit` runs this analysis and provides a table to identify; 
  * relation of each variable to nmds axis
  * how important a given variable is 
  * if the importance of each variable is significantly higher than would be expected by random
```{r general importance}
# Make labels for arrows and points (vectors and factors)
# vectorlabel1 <- colnames(dataset1)
# vectorlabel <- c(vectorlabel1[-1])
# label2 <- list("vectors" = vectorlabel,"factors"= label1)

#Generate table of importance values
importance <- envfit(nmds, dataset1, na.rm = TRUE)
str(nmds)
# #plot nmds
# # plot(nmds, dis="sites", type="n")
# # importance <- envfit(nmds, dataset1, na.rm = TRUE)
# plot(nmds, dis="sites", type="t")
# plot(nmds, dis="sites", type="n")
# orditorp(nmds, "sites", label = label1)
# plot(nmds)
# #plot(importance, labels = label2)
# 
# #Create biplot with importance vectors
# plot(importance, labels = label2)


# plot(nmds, dis="sites", type="t")
# plot(nmds, dis="sites", type="n")
# orditorp(nmds, "sites", label = label1)


# plot(importance, labels = label2)
# #generate a table of character importance
# importance <- envfit(nmds, dataset1, na.rm = TRUE)
# plot(nmds, "sites")


#  str(importance)
# class(importance$factors)



plot(nmds, dis="sites", type="t")
plot(nmds, dis="sites", type="n")
orditorp(nmds, "sites", label = label1)
plot(importance)

```

##Making a pretty dendrogram?

```{r}
# clus <- cutree(clust, 4)
# g <- split(names(clus), clus)
# 
# p <- ggtree(clust, linetype='dashed')
# clades <- sapply(g, function(n) MRCA(p, n))
# 
# p <- groupClade(p, clades, group_name='subtree') + aes(color=subtree)
# 
# # d <- data.frame(label = names(clus), 
# #                 cyl = datasetisla[names(clus), "cyl"])
# 
# ## Dummy grouping 
# d <- data.frame(label = names(clus), 
#                 pop = samplemeta$group1[match(names(clus),samplemeta$long.id)])
# 
# p %<+% d + 
#   layout_dendrogram() + 
#   geom_tippoint(aes(fill=factor(pop), x=x+.5),
#                 size=5, shape=21, color='black') +
#   geom_tiplab(aes(label=pop), size=3, hjust=.5, color='black') +
#   # geom_tiplab(angle=90, hjust=1, offset=-10, show.legend=FALSE) + 
#   scale_color_brewer(palette='Set1', breaks=1:4) +
#   theme_dendrogram(plot.margin=margin(6,6,80,6)) +
#   theme(legend.position=c(.9, .6))
```



##Mclust clustering analysis
```{r}
#cluster analysis- G forces dimensions of choice
#For Mclust to recognise an apriori group, that group must be represented in the NMDS by at least n = k+1 dimensions. So 3 samples inthis case. M. gibberagee only just satisfies this. 
mc <- Mclust(nmds$points)
plot(mc, what = "BIC")
plot(mc, what = "density")
plot(mc, what = "classification")
plot(mc, what = "uncertainty")

summary(mc)
#plot(mc)
 isladens <- densityMclust(nmds$points)
plot(isladens, what= "BIC")   #use the what command to run the plots individually without the number choice coming up each time
plot(isladens, what = "density", type = "hdr")



```
```{r testing character distributions by apriori groups}
#box and whisker plots. 



